---
output:
  html_document: default
  pdf_document: default
---
Marzena Malenta P4
========================================================

This is a project for Udacity Data Analyst Nanodegree which purpose is to conduct an analysis in Rstudio for the Loan Prosperity in United States of America. 
```{r echo=FALSE, message=FALSE, warning=FALSE, packages1}
chooseCRANmirror(graphics=FALSE, ind=1)
knitr::opts_chunk$set(echo = TRUE)
```



# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(ggthemes)
library(dplyr)
library(gridExtra)
library(tidyr)
library(MASS)
library(scales)
library(GGally)
library(RColorBrewer)
```


```{r echo=FALSE, Load_the_Data}
getwd()
setwd("C:/Users/Marcela/Desktop/Udacity/R/project/4")
# Load the Data
prosloan = read.csv("C:/Users/Marcela/Desktop/Udacity/R/project/4/prosperLoanData.csv")
```


getwd()
setwd("C:/Users/Marcela/Desktop/Udacity/R/project/4")
prosloan = read.csv("prosperLoanData.csv")

# Univariate Plots Section

Seems like the most common value to borrow is 4-5k, but we can observe that a range of 14-15k is also very popular

```{r echo=FALSE, Univariate_Plots1}
ggplot(aes(x = LoanOriginalAmount), data = prosloan) +
  geom_histogram(binwidth = 1000) +
  scale_x_continuous(
    limits = c(0, quantile(prosloan$StatedMonthlyIncome, 0.99,
                           na.rm = TRUE)),
    breaks = seq(0, quantile(prosloan$StatedMonthlyIncome, 0.99, 
                             na.rm = TRUE), 1000)) +
  theme(axis.text.x = element_text(angle = 90))
```
Loan Status: most of the loans are ongoing, but nearly a half of them is completed, then barely a quarter is changedoff. Surprisingy only small number of loans are past due.

```{r echo=FALSE, Univariate_Plots2}
  ggplot(data = prosloan, aes(x = LoanStatus)) +
  geom_bar(color="black", fill = '#007EE5') +
  ggtitle('Loan Status') +
  xlab('Loan Status') +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.6)) +
  ylab('Count')
```

Very interesting observation is that there is almost no difference in home properties belonging to the borrowers. We could have wrongly assume that borrowers might not be Homeowners due to financial problems, but the data assure us about this missconcepiton.

```{r echo=FALSE, Univariate_Plots3}
summary(prosloan$IsBorrowerHomeowner)  
```
Thought that owning a home in America tells a lot about your prosperity status and then necessity of taking a loan or mortgage, but borrowers of our dataset are divided almost perfectly in half with the ones that own a home or not.

```{r echo=FALSE, Univariate_Plots4}
ggplot(data = prosloan, aes(x = IsBorrowerHomeowner)) + 
  geom_bar(color="black", aes(fill = IsBorrowerHomeowner)) + 
  ggtitle('Is Borrower Home Owner') + 
  xlab('Is Borrower Home Owner') + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.6)) + 
  ylab('Count')+ 
  scale_fill_brewer(palette="Dark2")
```
6 Let's take a closer look at the salaries of borrwers, seems like most of our borrowers earn between 25-75k$ with the shape of a normal distribution.

```{r echo=FALSE, Univariate_Plots5}
ggplot(data = prosloan, aes(IncomeRange)) +
  geom_bar(color="black", fill = '#007EE5') +
  ggtitle('Borrower Income Range') +
  xlab('Income') +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.6)) +
  ylab('Count')
```

Let's take a look at Stated Monthly Income. It doesn't tell us much. 

```{r echo=FALSE, Univariate_Plots6}
ggplot(aes(x = StatedMonthlyIncome), data = prosloan) +
  geom_histogram(binwidth = 1000) +
  scale_x_continuous(
    limits = c(0, quantile(prosloan$StatedMonthlyIncome, 0.99,
                           na.rm = TRUE)),
    breaks = seq(0, quantile(prosloan$StatedMonthlyIncome, 0.99, 
                             na.rm = TRUE), 1000)) +
  theme(axis.text.x = element_text(angle = 90))
```
People earining 25.000 - 49.999$ tend to borrow more money.

```{r echo=FALSE, Univariate_Plots7}
ggplot(data = prosloan, aes(IncomeRange)) +
  geom_bar(color="black", fill = '#007EE5') +
  ggtitle('Borrower Income Range') +
  xlab('Income') +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.6)) +
  ylab('Count')
```


Prosper Score Distribution doesn't tell us much. We can only see that prosper scores 4,6,8 are the most common within borrowers. 

```{r echo=FALSE, Univariate_Plots8}
ggplot(data = prosloan, aes(x = ProsperScore)) + 
  geom_bar(color = "black", fill = '#007EE5') +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.6)) +
  xlab("Score") + ggtitle("Prosper Score Distribution")
summary(prosloan$ProsperScore)
```

Appearantly Americans are getting used to take credits, within last few years the amount of loans is getting strongly higher

```{r echo=FALSE, Load_the_Data9}
prosloan$ProsperRating <-factor(prosloan$ProsperRating..Alpha,
                               levels = c('AA', 'A', 'B', 'C', 'D', 'E', 'HR', 'NA'))
```


```{r echo=FALSE, Univariate_Plots10}
prosloan$LoanOriginationDate.month=format(prosloan$LoanOriginationDate, "(%m) %b")
months <- c('1', '2', '3', '4', '5', '6',
            '7', '8', '9', '10', '11', '12')
```


```{r echo=FALSE, Univariate_Plots11}
Sys.setlocale("LC_TIME", "C");
```
Very interesting observation is that in 2013 number of taken loans suddenly increased, this trend continued in the first quater of 2013.

```{r echo=FALSE, Univariate_Plots12}
ggplot(prosloan, aes(LoanOriginationDate.month)) +
  geom_histogram(stat="Count", color = 'black', fill = '#007EE5') +
  facet_wrap(~LoanOriginationDate.year) +
  ggtitle('Number of Loans by Month') +
  xlab('Month') +
  ylab('Number of Loans') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
```{r echo=FALSE, Load_the_Data13}
prosloan$ProsperRating <-factor(prosloan$ProsperRating..Alpha,
                               levels = c('AA', 'A', 'B', 'C', 'D', 'E', 'HR', 'NA'))
```


```{r echo=FALSE, Univariate_Plots14}
prosloan$ProsperRating.alpha = factor(prosloan$ProsperRating..Alpha.,
                                     levels = c("AA","A","B","C","D","E","HR","NA"))
```

It is very comprehensive that people with worse Prosper Rating need more money, so they borrow.
Graph below shows that the most loans are taken by people with Prosper Rating D and E.

```{r echo=FALSE, Univariate_Plots15}
ggplot(data = na.omit(prosloan), aes(ProsperRating)) +
  geom_bar(aes(fill = ProsperRating)) +
  ggtitle('Numbers of Loans by Prosper Rating') +
  xlab('Rating') +
  ylab('Number of Loans')
```
Monthly Loan Payments are very distinguished. We can observe that the most of loans have monthly payments of around 200$.

```{r echo=FALSE, Univariate_Plots16}
ggplot(data = prosloan, aes(x = MonthlyLoanPayment)) + 
  geom_bar(color = "black", fill = '#007EE5') +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.6)) +
  xlab("Score") + ggtitle("MonthlyLoanPayment")
```
Here we see that median of Monthly Loan Payment is 217.7 and the mean 272.5
```{r echo=FALSE, Univariate_Plots17}
summary(prosloan$MonthlyLoanPayment)
```

The most common amount to borrow is 4-5k$.

```{r echo=FALSE, Univariate_Plots18}
ggplot(aes(x = LoanOriginalAmount), data = prosloan) +
  geom_histogram(binwidth = 1000) +
  scale_x_continuous(
    limits = c(0, quantile(prosloan$StatedMonthlyIncome, 0.99,
                           na.rm = TRUE)),
    breaks = seq(0, quantile(prosloan$StatedMonthlyIncome, 0.99, 
                             na.rm = TRUE), 1000)) +
  theme(axis.text.x = element_text(angle = 90))
```

Loan Amounts divided by Interests and Fees:

```{r echo=FALSE, Univariate_Plots19}
ggplot(aes(x = LoanOriginalAmount), data = prosloan) +
  geom_histogram(binwidth = 1000) +
  scale_x_continuous(
    limits = c(0, quantile(prosloan$LP_InterestandFees, 0.99,
                           na.rm = TRUE)),
    breaks = seq(0, quantile(prosloan$LP_InterestandFees, 0.99, 
                             na.rm = TRUE), 1000)) +
  theme(axis.text.x = element_text(angle = 90))
```

```{r echo=FALSE, Univariate_Plots20}
prosloan$ProsperRating.alpha = factor(prosloan$ProsperRating..Alpha.,
                                     levels = c("AA","A","B","C","D","E","HR","NA"))
```

```{r echo=FALSE, Univariate_Plots21}
prosloan$LoanOriginationDate <- as.POSIXct(prosloan$LoanOriginationDate,
                                          format="%Y-%m-%d")
```

```{r echo=FALSE, Univariate_Plots22}
prosloan$LoanOriginationDate.year <-prosloan$LoanOriginationDate %>% 
  format('%Y') %>% strtoi()
```

  
summary(prosloan$ProsperRating.alpha)


# Univariate Analysis

### What is the structure of your dataset?

This report explores dataset containing information about xxx number loans 
There is 81 variables with 113937 observations.
We can simply categorize the dataset for borrowers and investors. The variables are divided by the information about borrowers and investors. It's like a closed circle, one drives another.
There would be no borrowers without investors etc.

### What is/are the main feature(s) of interest in your dataset?
The main features here are the two profiles represented in the dataset: borowerr and lender. The borrower part and lender part are charactarized by different sub points.
We can see what information use lendors about borrowers and vice versa.

### What other features in the dataset do you think will help support your \investigation into your feature(s) of interest?
The amount of loans and demogrpphical information.

### Did you create any new variables from existing variables in the dataset?
No, I didn't create any new variables. The ones provided were good enough.

### Of the features you investigated, were there any unusual distributions? \Did you perform any operations on the data to tidy, adjust, or change the form \of the data? If so, why did you do this?
No, the data was clean enoiugh to work on it.


# Bivariate Plots Section

Let's take a look at the Stated Monthly Income and Loan Amounts from this point of view:
```{r echo=FALSE, Bivariate_Plots1}
ggplot(data = prosloan, aes(x = StatedMonthlyIncome, y = LoanOriginalAmount)) +
  geom_point() +
  scale_x_discrete(breaks = seq(0, 203, 11))
```


Here we can see the Estimated Loss depends on Terms of Loans, apparently the biggest risk for investors brings 36 monhts loans, but it's only a slight difference between all of the terms.

```{r echo=FALSE, Bivariate_Plots2}
ggplot(aes(y = EstimatedLoss, x = Term), data = prosloan) +
  geom_boxplot()
```
Below we can see how correlated is the Term Loan to Estimated Return on the Investment for Lenders.
##26. Investor Profile - LenderYield ~ Term  
```{r echo=FALSE, Bivariate_Plots3}
ggplot(aes(y = EstimatedReturn, x = Term), data = prosloan) +
  geom_boxplot()
```

Taking a look at the correlation between Loan's Amount and Monthly Loan Payment we can logically see how they are depending one on another.
```{r echo=FALSE, Biivariate_Plots4}
ggplot(data = prosloan, aes(x = LoanOriginalAmount, y = MonthlyLoanPayment)) +
  geom_point() +
  scale_x_discrete(breaks = seq(0, 203, 11))
```

Looking at the plots of Stated Monthly Income and Loan ORiginal Amount we can't see nothing obvious
```{r echo=FALSE, Bivariate_Plots5}
ggplot(data = prosloan, aes(x = LoanOriginalAmount, y = StatedMonthlyIncome)) +
  geom_point() +
  scale_x_discrete(breaks = seq(0, 203, 11))
```

Income Range and Occupation:
```{r echo=FALSE, Biivariate_Plots6}
ggplot(data = prosloan, aes(x = Occupation, y = IncomeRange)) +
  geom_point() +
  scale_x_discrete(breaks = seq(0, 203, 11))
```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \investigation. How did the feature(s) of interest vary with other features in \the dataset?
The only relationship I found was the Amount of Loan and Monthly Payments, the higher the loan, the bigger monthly payment.

### Did you observe any interesting relationships between the other features \(not the main feature(s) of interest)?

No,I didn't observe any interesting relationship between features/interests.

### What was the strongest relationship you found?
The strongest relationship was between the Loan Original Amount and Monthly Payment. Logically the more you borrow, the more you pay back monthly.

# Multivariate Plots Section

Below we see Borrower Rates for Different Prosper Ratings, most of the boxes have similar sizes, we can see many outliers. 
```{r echo=FALSE, Multivariate_Plots1}
ggplot(data = prosloan, aes(x = Occupation, y = LoanOriginalAmount)) +
  geom_boxplot() +
  xlab("Prosper Rating") +
  ggtitle("Borrower Rate for Different Prosper Rating")
```
It's interesting that the Posper Rating B has most of 60 months term loans, while AA only 12.
```{r echo=FALSE, Multivariate_Plots2}
ggplot(aes( x = ProsperRating,y = Term, fill = LoanOriginalAmount),
       data = prosloan) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c('lightpink', 'yellow'))(100))
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \investigation. Were there features that strengthened each other in terms of \looking at your feature(s) of interest?

Heatmaps are a good tool to simply represnt this dataset. In my opinion it's more readable then tha charts and plots.

### Were there any interesting or surprising interactions between features?
No, nothing special.

### OPTIONAL: Did you create any models with your dataset? Discuss the \strengths and limitations of your model.
No, I didn't create any model.

------

# Final Plots and Summary


### Plot One

```{r echo=FALSE, Plot_One}
ggplot(data = prosloan, aes(x = EmploymentStatus, y = LoanOriginalAmount)) +
  geom_boxplot() +
  xlab("Prosper Rating") +
  ggtitle("Borrower Rate for Different Prosper Rating")
```

### Description One
The boxplot above shows clearly that the most ppl who receive loans are employed or self-employed. 

```{r echo=FALSE, Univariate_Plots88}
display.brewer.all()
brewer.pal(4, 'Set3')
```

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes( x = CreditGrade, y = IncomeRange, fill = LoanOriginalAmount),
       data = prosloan) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c('lightgreen', 'blue'))(100))
```
### Description Two
I chose this graph because in my opinion the Credit Grade is an essence in obtaining a loan at all.
This heatmap perfeclty shows that the better Credit Grade the more money you can borrow from the bank and that's the people with high Income Range obtain it.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes( x = EmploymentStatus,y = IncomeRange, fill = LoanOriginalAmount),
       data = prosloan) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c('lightblue', 'purple'))(100))
```

### Description Three
Another Heatmap which was important in my exploration is similar to the previous one but it takes into account Employment Status instead of Credit Grade. Curiously the "Other" emplyoment status has the biggest loans amount. 
------

# Reflection
When I saw this dataset I thought it would be way easier to conduct my analysis. I never worked before with R studio, so it was a challange, actually like the whole Nanodegree to me.
The struggle is that with 81 variables there is a lot of posibilites to draw plots and ask questions.
I couldn't find one question to answer or one field to focus that could give me a clear result.
We could look at the dataset from the point of view of the borrower or lender. I decided to taste both of them a little bit. At the beginning I thought that my previous project for Udacity using Tableau (you can find it here: https://public.tableau.com/profile/marzena5080#!/vizhome/MalaCaty1/Story2?publish=yes) would be helpful, but after using all the ideas that I had before, I realised that there is more needed.
And that's the essence of analyzing, you can discover many interesting things, in this project many of them led me nowhere and there was no clear relation between my results. 
Additional research could be necessary, like for example why in 2013 we observed a big peak in number of loans? Was it caused by economy growth or amendments in banks policy making it easier to obtain a credit? What pushed Americans to borrow money in that year?
Another insight is the Borrower State a part of dataset that could bring us many questions and answers, but I did not touch it here, I decided to not focus much on the demographical part which was well explained in my Tableau project. 
The only big connection which between lenders and borrowers is the Credit Grade and Emploment Status. In this closed circle bisuines, without a good job which leads to better Credit Grade obtaining a loan is not easy or sometimes even impossible. Banks lend money easier for people with profile. This is the essence of the loan business for both sides.  
I played with this dataset, not all graphs are logical and give a clear story, but it tought me a lot and at the end of my project I fell in love with heatmaps!
